\makeatletter
\def\sx@stop{\relax}\def\sx@open{(}\def\sx@close{)}
\def\sx@scan#1{%
  \def\sx@arg{#1}%        save the current token
  \ifx\sx@arg\sx@stop   % is it the end of the list?
    \let\sx@next\relax
  \else
    \ifx\sx@arg\sx@close\else\sx@space\fi % add space unless token is ")"
    {\sx@arg}%                            % write the token
    \ifx\sx@arg\sx@open                   
      \let\sx@space\relax                 % kill next space if token is "("
    \else
      \let\sx@space\;
    \fi
    \let\sx@next\sx@scan                  % scan next token
  \fi\sx@next
}
\newcommand{\sexp}[1]{\let\sx@space\relax\sx@scan#1\relax}
\makeatother

\title{Using genetic programming to evolve algorithms for audio synthesis}
\author{
Chris Donahue \\
Advisors: Peter Stone, Russell Pinkston \\
Department of Computer Science \\
University of Texas at Austin\vspace{-2ex}
}
\date{}

\documentclass[12pt]{article}

\usepackage{natbib}
\usepackage{verbatim}
\usepackage{amsmath}

\begin{document}
\maketitle

\begin{abstract}
This paper presents a system for evolving sound synthesis algorithms using the technique of genetic programming. The primary goal of this paper is to determine if genetic programming can be used to evolve sound synthesis algorithms whose output matches the timbre of a target sound. Typically this task involves many hours of human labor to determine an efficient way to mimic a given sound. An automation of this task would provide a useful, ``out of the box'' tool for matching target sounds and perhaps allow generalization to other center frequencies. This could be utilized as an enhanced sampling technique by electronic composers who wish to use a sound that they only have a recording of and not the instrument it was produced on.
\end{abstract}

\section{Timbre Matching Sound Synthesis Algorithms}

\subsection{Introduction}\label{INTRO}
Timbre is a collection of physical properties of a sound and is frequently explained in terms of its human utility rather than its underlying components. The American National Standards Institute defined timbre as ``that attribute of auditory sensation in terms of which a listener can judge that two sounds similarly presented and having the same loudness and pitch are dissimilar'' \citep{american1960american}. The ambiguity lies in determining what exactly ``that attribute'' is and what physical phenomena cause it. This psychoacoustic debate continues to this day; the advent of computer technology has introduced both a great road block and a great opportunity in our understand of timbre \citep{erickson1975sound}. Computers have the ability to render any sound representable by a discrete waveform and thus the umbrella term timbre can no longer be interpreted as a direct association between certain sounds and the concrete characteristics of the objects that produce them.

Early attempts to characterize the physical components of timbre mostly focused on the relative amplitudes and ratios of partial tones. These assumptions were based perhaps too heavily on Fourier's theorem that any periodic waveform can be decomposed into simple sin wave components \citep{helmholtz1857physiological, helmholtz1954sensations}. Like most aspects of music however, humans interpret not the absolute values of individual notes or frequencies but the way their relations change over time. Characterizations of timbre that place full emphasis on instantaneous spectral relationships are antiquated and have been augmented by multidimensional descriptions made up of several components with codependencies \citep{erickson1975sound}. One such description was proposed by J. F. Schouten in 1968 and involves five characteristics \citep{schouten1968perception, erickson1975sound}:
\begin{enumerate}
\item
The range between tonal and noise-like character
\item
The spectral envelope
\item
The time envelope in terms of rise, duration and decay
\item
The changes both of spectral envelope (formant-glide) and fundamental frequency (micro-intonation)
\item
The prefix, an onset of a sound quite dissimilar to the ensuing lasting vibration.
\end{enumerate}

For the purposes of this experiment, these five perceptual characteristics will be used to develop a method for comparing the timbral characteristics of two sounds. A computable metric of timbral comparison that is similar to human judgement is necessary for a genetic algorithm designed to produce synthesis algorithms capable of believably imitating target timbres.

\subsection{Motivations}
Chowning summarized the motivations for timbre matching research in his paper detailing the technique of frequency modulation audio synthesis.
\begin{quote}
Of interest in both acoustical re-search and electronic music is the synthesis of natural sound. For the researcher, it is the ultimate test of acoustical theory, while for the composer of electronic music it is an extraordinarily rich point of departure in the domain of timbre, or tone quality. \citep{chowning1973synthesis}
\end{quote}
Acoustic instruments are something of a rigor test for any new synthesis technique. They represent well-known ``problems'' that can be used to demonstrate the adaptability of any new algorithm and the scope of sounds that it can create. In the past, there have been many attempts to solve these problems by hand using a variety of synthesis techniques. This paper will present a system that allows one to supply a target sound and automatically receive a synthesis algorithm that mimics its timbral characteristics.

\subsection{Sound Synthesis Algorithms}\label{SSA}
Sound synthesis algorithms (SSA's) are typically numerical algorithms for rendering sonically interesting sounds \citep{tolonen1998evaluation}. Perhaps the quintessential SSA is the equation for rendering a sine wave. Mathematically, this equation can be expressed as $sin (2 \times \pi \times f \times t)$ where $f$ is the desired frequency of the sine wave and $t$ is the time since the start of the sine wave in seconds. In the digital domain, time at the $n$th sample would be $n/sample rate$.
	
The sine wave rendering equation could be expressed in several different ways. Using prefix notation, it can be expressed equivalently as the S-expression $\sexp{({sin} (\times (\times (\times 2 \pi) f) t))}$. S-expressions are computationally convenient as they allow easy evaluation and modification of an equation in a recursive tree structure. The evaluatable arrangements of operators and terminals (together referred to as primitives) induce a space of S-expressions. This concept will become more important when this paper details its use of genetic programming.
	
This sine wave rendering equation is so important to the domain of musical sounds that it will be convenient for our purposes to see it as a single, high level unit instead of the combination of eight lower level units as shown in FIGURE. Also included is a depiction of the sine wave rendering equation expressed as a more traditional synthesis topology. TODO: figures
	
\subsection{Mimicking Acoustic Instruments Using Sound Synthesis Algorithms}
Sound synthesis algorithms are a well studied aspect of computer music, and for as long as they have existed people have been using them to mimic recognizable acoustic timbres.

\subsubsection{Early Investigations (1950-1970)}
Max Mathews created the first widely used computer program for sound synthesis in 1957 called MUSIC. MUSIC and its descendants in the MUSIC-N family of programs utilize wrapper functions around common synthesis routines called unit generators. This is a similar concept to wrapping the sine wave rendering equation into a single S-expression primitive as discussed in \ref{SSA}. Mathews wrote that ``there are no theoretical limitations to the performance of the computer as a source of musical sounds, in contrast to the performance of ordinary instruments.'' \citep{mathews01111963} While this statement can be challenged due to a computer's lack of ability to produce continuous signals, it holds perceptually as any discrete wave form can be rendered by a computer. Jean-Claude Risset worked with Mathews in the late 1960's to attempt to produce with a computer sounds that mimicked acoustic instruments. While they were successful in creating brass-like and string-like sounds by analyzing the spectral content of recordings, their conclusion was that ``additional factors must be discovered to yield an impeccable match to a given [instrument]'' \citep{risset1969analysis}.

\subsubsection{FM Synthesis (1970-1990)}
In 1973 John Chowning introduced a method for utilizing frequency modulation to efficiently synthesize sounds with rich harmonic spectra that developed over time. He states in the introduction of the related paper that ``temporal evolution of the spectral components is of critical importance in the determination of timbre'' \citep{chowning1973synthesis}. The task of efficiently producing spectrally complex tones had presented a major hurdle to mimicking spectrally complex acoustic instruments as computer performance was limited. Chowning himself used FM synthesis to mimic brass, woodwind, percussive sounds, and later a singing voice \citep{chowning1989frequency} with impressive subjective results. Perhaps one of the ``additional factors'' elusive to Mathews and Risset was the temporal evolution of spectral components that FM synthesis provides.
Work in the 1980's on mimicking acoustic timbres was largely commercial after Stanford licensed Chowning's work on FM audio synthesis to Yamaha. Audio researchers at Yamaha continued to mimic tones of acoustic instruments by modifying the parameters and carrier/modulation structure of FM synthesis.

\subsubsection{Genetic Algorithms (1990-present)}
\paragraph{Traditional Genetic Algorithms}
Genetic algorithms (GA's) were only popularized in the 1980's despite previous decades of theoretical study. TODO: EXPLAIN GENETIC ALGORITHM

In 1993, Horner demonstrated a system to automate timbre matching using a genetic algorithm \citep{horner1993machine}. Horner successfully demonstrated a system for optimizing the parameters of a fixed FM synthesis structure with a single modulator and multiple parallel carriers. Johnson used an interactive genetic algorithm to control parameters for a granular synthesis algorithm \citep{johnson1999exploring} and Wehn used ideas from natural selection to explore the parameter space and in a more limited sense the connection space of FM-like synthesizers \citep{wehn1998using}. Genetic algorithms are well-suited for searching large, multi-dimensional spaces such as the parameters of well-known synthesis algorithms and are a natural choice for such a task.

\paragraph{Genetic Programming}
John Koza's 1994 book defined a new class of genetic algorithms altogether \citep{koza1992genetic}. Koza borrowed concepts from traditional genetic algorithms to evolve computer programs and called the technique genetic programming (GP). The computer programs that GP evolves are represented as S-expressions. An S-expression is a tree representation of a program which can be anything from a simple mathematical expression to much more complicated algorithms involving logic operators and domain-specific operations. TODO: EXPLAIN GENETIC PROGRAMMING
			
The domain-specific applications of genetic programming are intriguing for sound synthesis as nodes of the S-expression could represent functional units such as sin wave oscillators and time-varying filters as discussed previously. In 2000, Ricardo Garcia successfully demonstrated the technique of GP in the domain of target-matching sound synthesis algorithms \citep{garcia2000towards}.

Garcia's approach evolved trees that represented a set of instructions for building sound synthesis topologies such as the topology depicted in figure (TODO: FIGURE ABOVE FROM \ref{SSA}). This indirect encoding was motivated by the fact that sound synthesis topologies can be cyclical and S-expressions cannot be. However, these non-cyclical ``instruction trees'' could contain instructions to build a cyclical synthesis topology and therefore allow unmodified genetic programming to be utilized to create cyclical synthesis algorithms.

Garcia made the following statement about his instruction trees ``every [instruction] tree maps into a single topology... but a topology can be represented by many different [instruction] trees.'' \citep{garcia2012automatic}. This leads me to believe that this approach is perhaps not the ideal way to apply GP to sound synthesis as this representation is searching through a larger space of S-expressions that map into a smaller space of synthesis algorithms. No examples of hand-designed SSA's that utilize cyclic topologies to mimic acoustic instruments were found in the research for this paper which also suggests that the cost of search efficiency is not worth this feature. Regardless, cyclic behavior could also be represented non-cyclically by the use of a delay buffer primitive if it is found to be necessary. Additionally, Garcia uses sub-optimization of a known number of constants once a topology had been produced. This is an interesting strategy but could theoretically be encoded into GP itself by using ephemeral random constants with numeric mutation \citep{evett1998numeric}.

\subsection{Target-Matching Genetic Programming Search for Sound Synthesis Algorithms}
\subsubsection{Method}
This experiment will use genetic programming to evolve synthesis algorithms directly, utilizing a set of primitives identified as essential components for mimicking timbres. The inputs to this system are a recorded target sound represented by a WAV file and the fundamental frequency of the target sound. This method differs from Garcia's whose representation did not involve the fundamental frequency as input. I argue that providing the fundamental frequency is a good addition to the inputs of the system as it prevents the algorithm from having to waste search time to find the fundamental and allows the experimenter to test how well the discovered algorithm generalizes to other pitches.
\subsubsection{Available primitives}
Arity is the number of input arguments a particular primitive requires. Primitives with arity of $0$ are the terminal nodes of S-expressions. These are the primitives used in all tests unless otherwise noted. TODO: add/remove primatives found to be necessary/not useful
\begin{center}
\begin{tabular}{ | l | l | } \hline
Type & Arity \\ \hline
Addition & 2 \\ \hline
Multiplication & 2 \\ \hline
Frequency modulation oscillator & 2 \\ \hline
Time-varying filters & 3 \\ \hline
ADSR envelope & 1 \\ \hline
Sin wave oscillator & 0 \\ \hline
White noise & 0 \\ \hline
Ephemeral random constants & 0 \\ \hline
\end{tabular}
\end{center}
TODO: params
\subsubsection{Loop}
The process of this experiment will involve the following main loop:
\begin{enumerate}
\item
Initial sound synthesis algorithm population
\item
LOOP START:
Run each algorithm to render sound
\item
Evaluate the sound in comparison to the target by means of a fitness function
\item
If $best\ fitness\ found\ <\ threshold\ ||\ number of generations\ >=\ limit$, FINISH and return champion
\item
otherwise render the next generation of individuals using traditional genetic programming operations and return to LOOP START
\end{enumerate}
TODO: params
\subsubsection{Initial Population}
The random generation of the initial population is a mostly unmodified implementation of Koza's ramped half and half method used to generate diversity in the initial population \citep{koza1992genetic}. The only difference is that I do not check for duplicate S-expressions. The performance gain from not evaluating the same S-expression more than once does not seem to be worth the performance hit of comparing each new S-expression to all existing ones. I am using the technique of ephemeral random constants to propagate random values throughout the initial population.
TODO: params
\subsubsection{Fitness Function}
The fitness function used for these tests was designed with the five parameters listed in \ref{INTRO} in mind. The target samples are partitioned into $m$ groups of $n$ samples with $o$ overlap between groups. These groups are analyzed using a discrete Fourier transform which can be used to determine magnitude and phase at $n/2$ frequency bins equally spaced between $0$ and $\frac{sample rate}{2}$ $Hz$.

The following values are calculated for each bin $j$ of group $i$:
\begin{itemize}
\addtolength{\itemindent}{1cm}
\vspace{-3mm}
\item $EMAC_{(i, j)}$: Exponential moving average value at the bin
\vspace{-3mm}
\item $TMAG_{(i, j)}$: Magnitude of the bin
\vspace{-3mm}
\item $TPHS_{(i, j)}$: Phase of the bin
\vspace{-3mm}
\item $UNDER_{(i, j)}$: Penalty for candidate undershooting the bin
\vspace{-3mm}
\item $OVER_{(i, j)}$: Penalty for candidate overshooting the bin
\vspace{-2mm}
\end{itemize}

The following values are calculated for each group $i$:
\begin{itemize}
\addtolength{\itemindent}{1cm}
\vspace{-3mm} 
\item $MAXABOVE_{(i)}$: $\max{(MAG_{(i, j)} - EMAC_{(i, j)})}$
\vspace{-3mm} 
\item $MAXBELOW_{(i)}$: $\max{(EMAC_{(i, j)} - MAG_{(i, j)})}$
\vspace{-2mm}
\end{itemize}

$UNDER_{(i, j)}$ and $OVER_{(i, j)}$ are calculated for each bin $j$ of group $i$ in the following way:
\vspace{-3mm} 
\begin{align*}
if (TMAG_{(i, j)} \le EMAC_{(i, j)}): \\
UNDER_{(i, j)} &= \frac{EMAC_{(i, j)} - TMAG_{(i, j)}}{MAXBELOW_{(i)}} * good + base \\
OVER_{(i, j)} &= \frac{EMAC_{(i, j)} - TMAG_{(i, j)}}{MAXBELOW_{(i)}} * bad + base \\
if (TMAG_{(i, j)} > EMAC_{(i, j)}): \\
UNDER_{(i, j)} &= \frac{TMAG_{(i, j)} - EMAC_{(i, j)}}{MAXABOVE_{(i)}} * bad + base \\
OVER_{(i, j)} &= \frac{TMAG_{(i, j)} - EMAC_{(i, j)}}{MAXABOVE_{(i)}} * good + base
\end{align*}
\vspace{-2mm}

Each candidate S-expression is rendered to produce a candidate wave form with the same number of samples as the target. The candidate samples are partitioned into groups of $n$ with $o$ overlap in the same way as the target samples.

The following values are calculated for each bin $j$ of group $i$ from the candidate samples:
\begin{itemize}
\addtolength{\itemindent}{1cm}
\vspace{-3mm} 
\item $CMAG_{(i, j)}$: Magnitude of the frequency bin
\vspace{-3mm} 
\item $CPHS_{(i, j)}$: Phase of the frequency bin
\vspace{-2mm}
\end{itemize}

Each candidate is assigned fitness in the following way:
\vspace{-2mm} 
\begin{align*}
PHSERROR_{(i, j)} &= (|TPHS_{(i, j)} - CPHS_{(i, j)}|)^{PHASEPENALTY} \\
if (CMAG_{(i, j)} \le TMAG_{(i, j)}): \\
MAGERROR_{(i, j)} &= (TMAG_{(i, j)} - CMAG_{(i, j)})^{UNDER_{(i, j)}} \\
if (CMAG_{(i, j)} > TMAG_{(i, j)}): \\
MAGERROR_{(i, j)} &= (CMAG_{(i, j)} - TMAG_{(i, j)})^{OVER_{(i, j)}}
\end{align*}
\vspace{-8mm}
\begin{align*}
FITNESS &= PHASEWEIGHT * (\sum_{i=1}^{m} \sum_{j=1}^{\frac{n}{2} - 1} PHSERROR_{(i, j)}) \\
        &+ MAGWEIGHT * (\sum_{i=1}^{m} \sum_{j=1}^{\frac{n}{2} - 1} MAGERROR_{(i, j)})
\end{align*}
\vspace{-4mm}

The motivations for this fitness function come from the attributes of timbre listed in \ref{INTRO}. At a high level, the weighting system is giving a higher penalty for a candidate overshooting a target valley and a higher penalty for a candidate undershooting a target peak. Penalty scaling is constant for a candidate undershooting a target valley and overshooting a target peak. This has the two-pronged effect of ensuring that better candidates in the early generations emphasize spectral peaks and don't try to match noisy bins. Later in the generations when the population begins to converge, the GA can continue to optimize by reducing the error at peaks and filling in lower magnitude spectral content.

This fitness function was empirically found to produce spectral content that roughly matched the sustain portion of targets but was worse at matching overall amplitude envelope. Experiments were performed where groups were weighted by their magnitude average's difference from the average magnitude across all groups. This placed higher fitness emphasis on transient groups, where the amplitude spectrum was the most different from the average (usually the attack/decay periods of the target). However, this technique was empirically found to lower the quality of the spectral matching.

TODO: params
\subsubsection{Genetic Operations}
The utilized genetic operations are those typically associated with genetic programming. Individuals are selected proportionally to their assigned fitnesses (where lower is better due to the fact that fitness is a calculation of error).
\begin{itemize}
\item
Crossover: Select two parents, swapping two randomly selected subtrees from within them and adding both resultant trees to the next generation. This operation is quite expressive as the heights of the resultant children can range from 0 to the sum of the heights of the parents. It can produce radically different children even if the parents are the same
\item
Reproduction: Select one algorithm and clone it without mutation into the next generation.
\item
Mutation: Select one algorithm and select one node $s$ from it uniformly at random. Replace this node with a new subtree with the $Grow$ method and max height of $maxheight - s.depth$.
\item
Numerical mutation: Select one algorithm and numerically mutate all of its coefficients.

TODO: params/include figures from Koza/elaborate on numerical mutation
\end{itemize}
\subsubsection{Experiments}
\begin{comment}
Ideally one would ``clean'' the WAV file as much as possible by removing areas of silence, removing any DC offset and normalizing.
\end{comment}
\subsubsection{Results}
\subsubsection{Identified Limitations}
Because of the nature of S-expressions, a given node cannot have multiple parents. This theoretically does not limit the capabilities of the sound synthesis algorithms produced by GP because an S-expression can have multiple identical terminals that would be identical in behavior to a node with multiple parents. However, this greatly increases search complexity and computation time to match discover efficient algorithms such as FM synthesis using a single modulator and multiple carriers (the topology that \citep{horner1993machine} optimized using a genetic algorithm). Another limitation is lack of support for cyclic algorithms although as was hypothesized earlier this might not be of too much concern in the domain of target matching.
\subsubsection{Conclusions}
\subsubsection{Parameters}

\begin{comment}

\section{Interactive Evolution of Novel Sound Synthesis Algorithms}
\subsection{Motivation}
\begin{quote}
\emph{To seek out new tonalities, new timbres... \\
To boldly listen to what no one has heard before.} \citep{sethares2004tuning}
\end{quote}
\begin{quote}
The musical value of the computer does not lie, of course, in its ability to duplicate exactly what a real instrument can do, but rather in yielding an extended repertory of sounds, including and going beyond the classes of sounds of actual instruments. \citep{risset1969analysis}.
\end{quote}

The concept of timbre is largely compartmentalized by modern composers. Acoustic composers typically restrict themselves to given sounds, such as the sounds that instruments in a wind ensemble or orchestra can produce. Electronic and electro-acoustic composers have a wider timbral palette to work with but usually restrict themselves to working within the confines of available synthesizers and samples. It has a hard task to discover novel sounds from thoroughly investigated synthesis techniques. Wishart said in his book Audible Design \citep{wishart1994audible}:
\begin{quote}
The spectral characteristics of sound have, for so long, been inaccessible to the composer that we have become accustomed to lumping together all aspect of the spectral structure under the catch all term ``timbre''...
\end{quote}	 
These issues suggest the need for alternative and accessible methods of timbre investigation for musicians and researchers. Such capabilities could theoretically allow a musician to utilize an instrument they do not have physical access too from a single recording of its sound. Many electronic musicians utilize synthesis techniques such as additive, subtractive, ring modulation, amplitude modulation, and frequency modulation to create sounds. This paper will also present an interactive genetic algorithm for searching the space of sound synthesis algorithms of which these traditional synthesis techniques are a minuscule subset of the search space.

\subsection{Experimental Setup}
\subsubsection{Method}
This experiment will use genetic programming to evolve synthesis algorithms directly, utilizing any set of primitives that the user chooses. The inputs to this system will be the user-selected primitives and the output will be multiple SSA's for the user to modify and evaluate. The user will be able to step through the evolutionary process, assigning fitness and auditioning the produced SSA's using a MIDI keyboard or virtual MIDI keyboard. The user may manipulate any variable constants of the produced SSA. The user may opt to input an audio file and let the synthesizer attempt to evolve towards it. The user may also let evolution experiment randomly using novelty search with unencountered fundamental/partial frequency ratios as the novelty goal.
\subsubsection{Results}
\subsubsection{Identified Limitations}
\subsubsection{Conclusions}
\subsubsection{Parameters}

\end{comment}

\section{Implementation}
An early prototype for the interactive system was built using Cabbage, Csound and Python. After several performance limitations with this setup were encountered the implementation was switched to C++ and utilized the JUCE audio/GUI library. JUCE is a popular open-source library for developing audio plug-ins and was used as the core audio engine of the implementation. This project could not have been possible without JUCE and author/maintainer Jules's relentless demand for quality. Filter primitives were developed by modifying Tony Fisher's filter design C program. Spectrum analysis for the fitness function was performed by Kiss FFT, a relatively fast and extremely lightweight implementation of the Fast Fourier transform.

Target-matching tests were run on the UT Condor Computing Cluster. \begin{comment}Interactive evolution tests were run on a variety of different machines. TODO: fill in with actual data \end{comment}

\section{Future Work}
The system will also allow the user to examine, test and adjust these discovered algorithms to improve their fit.

\section{Acknowledgements}
The author would like to thank the Computational Intelligence in Game Design lab, part of the University of Texas's Freshman Research Initiative program, for providing the tools and workspace to implement this project. Thanks to Joel Lehman for advice on the genetic algorithm aspects of this project. Thanks to Ashvin Bashyam for his advice on the subtleties of working in a frequency domain representation of a signal. 
\begin{comment}
Thanks to TESTERS for kindly agreeing to test the interactive genetic algorithm.
\end{comment}
Finally, the author wishes to express gratitude to his advisors and committee members who were extremely helpful throughout the process.

%\section{For More Information and Download}
%cs.utexas.edu/~cdonahue/GPSynth.html (audio samples)
%github.com/cdonahue/gpsynth (source and downloadable/plugin versions of the interactive GA)

%\paragraph{Outline}
%The remainder of this article is organized as follows.
%Section~\ref{previous work} gives account of previous work.
%Our new and exciting results are described in Section~\ref{results}.
%Finally, Section~\ref{conclusions} gives the conclusions.

%\section{Results}\label{results}
%In this section we describe the results.

\bibliographystyle{abbrv}
\bibliography{bibthesis}

\end{document}
